---
title: "Introduction to visNetwork"
author: "B. Thieurmel - DataStorm"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to visNetwork}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**visNetwork** is a R package for network visualization, using **vis.js** javascript library (http://visjs.org). All the remarks and bugs returns are welcome on github : https://github.com/datastorm-open/visNetwork.

## Minimal example

**visNetwork** needs at least two informations :

* a nodes data.frame, with *id* column
* a edges data.frame, with *from* and *to* columns

```{r}
require(visNetwork, quietly = TRUE)
# minimal example
nodes <- data.frame(id = 1:3)
edges <- data.frame(from = c(1,2), to = c(1,3))
visNetwork(nodes, edges, width = "100%")
```

## Find help

Besides the help R functions, a vignette is available, and you can access and read the full javascript API :

```{r, eval = FALSE}
visDocumentation()
vignette("Introduction-to-visNetwork") # with CRAN version
# shiny ?
shiny::runApp(system.file("shiny", package = "visNetwork"))
```

Or visit our online help page : http://datastorm-open.github.io/visNetwork/.

## Simple individual nodes and edges customization

* Adding more variables on nodes data.frame. See **_visNodes_** for available options.

```{r}
nodes <- data.frame(id = 1:10, 
                    label = paste("Node", 1:10),                                 # add labels on nodes
                    group = c("GrA", "GrB"),                                     # add groups on nodes 
                    value = 1:10,                                                # size adding value
                    shape = c("square", "triangle", "box", "circle", "dot", "star",
                    "ellipse", "database", "text", "diamond"),                   # control shape of nodes
                    title = paste0("<p><b>", 1:10,"</b><br>Node !</p>"),         # tooltip (html or character)
                    color = c("darkred", "grey", "orange", "darkblue", "purple"),# color
                    shadow = c(FALSE, TRUE, FALSE, TRUE, TRUE))                  # shadow

head(nodes)
```

* Adding more variables on edges data.frame. See **_visEdges_** for available options.

```{r}
edges <- data.frame(from = sample(1:10, 8), to = sample(1:10, 8),
                    label = paste("Edge", 1:8),                                 # add labels on edges
                    length = c(100,500),                                        # length
                    arrows = c("to", "from", "middle", "middle;to"),            # arrows
                    dashes = c(TRUE, FALSE),                                    # dashes
                    title = paste("Edge", 1:8),                                 # tooltip (html or character)
                    smooth = c(FALSE, TRUE),                                    # smooth
                    shadow = c(FALSE, TRUE, FALSE, TRUE))                       # shadow
head(edges)
```

```{r}
visNetwork(nodes, edges, width = "100%")
```

## Global nodes/edges configuration

* Set global options for nodes and edges using **_visNodes_** and **_visEdges_**, and use options per group using **_visGroups_**.

```{r}
nodes <- data.frame(id = 1:5, group = c(rep("A", 2), rep("B", 3)))
edges <- data.frame(from = c(2,5,3,3), to = c(1,2,4,2))

visNetwork(nodes, edges, width = "100%") %>% 
  visNodes(shape = "square") %>%                        # square for all nodes
  visEdges(arrows ="to") %>%                            # arrow "to" for all edges
  visGroups(groupname = "A", color = "darkblue") %>%    # darkblue for group "A"
  visGroups(groupname = "B", color = "red")             # red for group "B"
```

# Network configuration
Configuration options are available in *visOptions*, *visInteraction*, *visLayout*, *visHierarchicalLayout*, *visPhysics* :

## Example Data
```{r, echo=TRUE}
nb <- 10
nodes <- data.frame(id = 1:nb, label = paste("Label", 1:nb),
 group = sample(LETTERS[1:3], nb, replace = TRUE), value = 1:nb,
 title = paste0("<p>", 1:nb,"<br>Tooltip !</p>"), stringsAsFactors = FALSE)

edges <- data.frame(from = trunc(runif(nb)*(nb-1))+1,
 to = trunc(runif(nb)*(nb-1))+1,
 value = rnorm(nb, 10), label = paste("Edge", 1:nb),
 title = paste0("<p>", 1:nb,"<br>Edge Tooltip !</p>"))
```

## Add legend
New feature in __visNetwork_0.1.2__ : you can now add more custom legend on nodes / edges ! Default on groups (like in previous versions) : 

```{r}
visNetwork(nodes, edges, width = "100%") %>% visLegend()
```

Or passing data : 

```{r}
visNetwork(nodes, edges, width = "100%") %>% 
  visLegend(useGroups = FALSE, addNodes = data.frame(label = "Nodes", shape = "circle"), 
            addEdges = data.frame(label = "link", color = "black"))
```

## Highlight nearest
You can highlight nearest nodes and edges clicking on a node with **highlightNearest** option :

```{r}
visNetwork(nodes, edges, width = "100%") %>% 
  visOptions(highlightNearest = TRUE)
```

It's now possible to control the degree of depth (visNetwork >= 0.1.2) : 

```{r}
visNetwork(nodes, edges, width = "100%") %>% 
  visOptions(highlightNearest = list(enabled =TRUE, degree = 2))
```


## Select by node id
You can also select nodes by id/label with a list  with **nodesIdSelection** : 

```{r}
visNetwork(nodes, edges, width = "100%") %>% 
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE)
```

## Select by a column
New feature in __visNetwork_0.1.2__ : select some nodes by the values of a column using **selectedBy** option :

```{r}
# can be the column you want
nodes$sel <- sample(c("sel1", "sel2"), nrow(nodes), replace = TRUE)
visNetwork(nodes, edges, width = "100%") %>%
 visOptions(selectedBy = "sel")
```

## Directed Network
```{r, echo = FALSE}
nodes <- data.frame(id = 1:nb, label = paste("Label", 1:nb),
 group = sample(1:nb, nb, replace = TRUE), value = 1:nb,
 title = paste0("<p>", 1:nb,"<br>Tooltip !</p>"), stringsAsFactors = FALSE)

edges <- data.frame(from = trunc(runif(nb)*(nb-1))+1,
 to = trunc(runif(nb)*(nb-1))+1,
 title = paste0("<p>", 1:nb,"<br>Edge Tooltip !</p>"))
```

```{r}
visNetwork(nodes, edges, width = "100%") %>% 
  visEdges(arrows = 'from')
```

## Custom navigation
*vis.js* propose some navigation tools :

```{r}
visNetwork(nodes, edges, width = "100%") %>% 
  visInteraction(navigationButtons = TRUE)
```

## Data Manipulation
And some data manipulation tools :

```{r}
visNetwork(nodes, edges, width = "100%") %>% 
  visOptions(manipulation = TRUE)
```

## Hierarchical Layout
You can use and control hierarchical layout with *visHierarchicalLayout* and *visLayout* :

```{r, echo = TRUE}
nodes <- data.frame(id = 1:7)

edges <- data.frame(
  from = c(1,2,2,2,3,3),
  to = c(2,3,4,5,6,7)
)
```

```{r}
visNetwork(nodes, edges, width = "100%") %>% 
  visEdges(arrows = "from") %>% 
  visHierarchicalLayout() 
# same as   visLayout(hierarchical = TRUE) 

visNetwork(nodes, edges, width = "100%") %>% 
  visEdges(arrows = "from") %>% 
  visHierarchicalLayout(direction = "LR")
```

## Freeze network
```{r}
visNetwork(nodes, edges, width = "100%") %>% 
  visInteraction(dragNodes = FALSE, dragView = FALSE, zoomView = FALSE)
```

## Use font awesome icons in your network

You can use **Font Awesome** icons using groups or nodes options. **Font Awesome** library (http://fortawesome.github.io/Font-Awesome/) is not part of dependencies. use addFontAwesome() if needed.

```{r}
nodes <- data.frame(id = 1:3, group = c("B", "A", "B"))
edges <- data.frame(from = c(1,2), to = c(2,3))

visNetwork(nodes, edges, width = "100%") %>%
  visGroups(groupname = "A", shape = "icon", icon = list(code = "f0c0", size = 75)) %>%
  visGroups(groupname = "B", shape = "icon", icon = list(code = "f007", color = "red")) %>%
  visLegend() %>%
  addFontAwesome()
```

# Additional features

## Visualize rpart object

New *visTree* function allows to visualize and customize a *rpart* classification and regression tree. Have a look to *visTreeEditor* to edity and get back network, or to *visTreeModuleServer* to use custom tree module in R.

```{r, eval = T}
library(rpart)
# Complex tree
data("solder")
res <- rpart(Opening~., data = solder, control = rpart.control(cp = 0.00005))
visTree(res, height = "800px", nodesPopSize = TRUE, minNodeSize = 10, maxNodeSize = 30)

```

## Use in Shiny

With *visNetworkOutput* and *renderVisNetwork*. Using with shiny, and enabled options *nodesIdSelection*, you can access to new input with current selection value. Morevoer, you can do a lot of things with *visNetworkProxy*

```{r, eval = FALSE}
output$mynetwork <- renderVisNetwork({... visOptions(nodesIdSelection = TRUE)}) # created input$mynetwork_selected

```

And with *selectedBy*, new input __input$mynetwork_selectedBy__.

## Physics, tooltip, events, 

You can also control :

* physics of network : *visPhysics*
* events : *visEvents*

## Export

```{r, eval = FALSE}
network <- visNetwork(nodes, edges, width = "100%")
visSave(network, file = "network.html")
```

## Use DOT language data 

```{r}
visNetwork(dot = 'dinetwork {1 -> 1 -> 2; 2 -> 3; 2 -- 4; 2 -> 1 }', width = "100%")
```

## Use gephi json export file

```{r, eval = FALSE}
# don't run here
visNetwork(gephi = 'WorldCup2014.json')
```

