---
title: "Aesthetic specifications"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aesthetic specifications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(ggplot2)
knitr::opts_chunk$set(fig.dpi = 96)
```

This vignette summarises the various formats that grid drawing functions take.  Most of this information is available scattered throughout the R documentation.  This appendix brings it all together in one place. 

## Colour

Colours can be specified with: 

*   A __name__, e.g., `"red"`. R has `r length(colours())` built-in named
    colours, which can be listed with `colours()`. The Stowers Institute 
    provides a nice printable pdf that lists all colours: 
    <http://research.stowers-institute.org/efg/R/Color/Chart/>.

*   An __rgb specification__, with a string of the form `"#RRGGBB"` where each of 
    the pairs `RR`, `GG`, `BB` consists of two hexadecimal digits giving a value 
    in the range `00` to `FF`
    
    You can optionally make the colour transparent by using the form 
    `"#RRGGBBAA"`.

*   An __NA__, for a completely transparent colour.

*   The [munsell](https://github.com/cwickham/munsell) package, by Charlotte 
    Wickham, provides a wrapper around the colour system designed by Alfred 
    Munsell.

## Line type {#sec:line-type-spec}

Line types can be specified with:

*   An __integer__ or __name__: 0 = blank, 1 = solid, 2 = dashed, 3 = dotted, 
    4 = dotdash, 5 = longdash, 6 = twodash, as shown below:

    ```{r}
    lty <- c("blank", "solid", "dashed", "dotted", "dotdash", 
         "longdash","twodash")
    linetypes <- data.frame(
      y = seq_along(lty),
      lty = lty
    ) 
    ggplot(linetypes, aes(0, y)) + 
      geom_segment(aes(xend = 5, yend = y, linetype = lty)) + 
      scale_linetype_identity() + 
      geom_text(aes(label = lty), hjust = 0, nudge_y = 0.2) +
      scale_x_continuous(NULL, breaks = NULL) + 
      scale_y_continuous(NULL, breaks = NULL)
    ```

*   The lengths of on/off stretches of line. This is done with a string 
    containing 2, 4, 6, or 8 hexadecimal digits which give the lengths of
    consecutive lengths. For example, the string `"33"` specifies three units 
    on followed by three off and `"3313"` specifies three units on followed by 
    three off followed by one on and finally three off. 
    
    The five standard dash-dot line types described above correspond to 44, 13, 
    1343, 73, and 2262.

The `size` of a line is its width in mm.

## Shape {#sec:shape-spec}

Shapes take four types of values:

*   An __integer__ in $[0, 25]$:

    ```{r}
    shapes <- data.frame(
      shape = c(0:19, 22, 21, 24, 23, 20),
      x = 0:24 %/% 5,
      y = -(0:24 %% 5)
    )
    ggplot(shapes, aes(x, y)) + 
      geom_point(aes(shape = shape), size = 5, fill = "red") +
      geom_text(aes(label = shape), hjust = 0, nudge_x = 0.15) +
      scale_shape_identity() +
      expand_limits(x = 4.1) +
      scale_x_continuous(NULL, breaks = NULL) + 
      scale_y_continuous(NULL, breaks = NULL)
    ```

*   A __single character__, to use that character as a plotting symbol.

*   A `.` to draw the smallest rectangle that is visible, usualy 1 pixel.
   
*   An `NA`, to draw nothing.

Note that shapes 21-24 have both stroke `colour` and a `fill`. The size of the filled part is controlled by `size`, the size of the stroke is controlled by `stroke`. Each is measured in mm, and the total size of the point is the sum of the two. Note that the size is constant along the diagonal in the following figure.

```{r}
sizes <- expand.grid(size = (0:3) * 2, stroke = (0:3) * 2)
ggplot(sizes, aes(size, stroke, size = size, stroke = stroke)) + 
  geom_abline(slope = -1, intercept = 6, colour = "white", size = 6) + 
  geom_point(shape = 21, fill = "red") +
  scale_size_identity()
```


## Text

### Font size


### Font face

There are only three fonts that are guaranteed to work everywhere: "sans" (the default), "serif", or "mono":

```{r}
df <- data.frame(x = 1, y = 3:1, family = c("sans", "serif", "mono"))
ggplot(df, aes(x, y)) + 
  geom_text(aes(label = family, family = family))
```

It's trickier to include a system font on a plot because text drawing is done differently by each graphics device (GD). There are five GDs in common use (`png()`, `pdf()`, on screen devices for Windows, Mac and Linux), so to have a font work everywhere you need to configure five devices in five different ways. Two packages simplify the quandary a bit: 

* `showtext` makes GD-independent plots by rendering all text as polygons. 

* `extrafont` converts fonts to a standard format that all devices can use. 

Both approaches have pros and cons, so you will to need to try both of them and see which works best for your needs.

### Family

<!--
postscriptFonts, pdfFonts, quartzFonts

Find R news article

\begin{itemize}
  \item \code{face}
  \item \code{family}
  \item \code{lineheight}
  \item \code{fontsize}
\end{itemize}
-->


### Justification

Horizontal and vertical justification have the same parameterisation, either a string ("top", "middle", "bottom", "left", "center", "right") or a number between 0 and 1:

* top = 1, middle = 0.5, bottom = 0
* left = 0, center = 0.5, right = 1


```{r}
just <- expand.grid(hjust = c(0, 0.5, 1), vjust = c(0, 0.5, 1))
just$label <- paste0(just$hjust, ", ", just$vjust)

ggplot(just, aes(hjust, vjust)) +
  geom_point(colour = "grey70", size = 5) + 
  geom_text(aes(label = label, hjust = hjust, vjust = vjust))
```

Note that you can use numbers outside the range (0, 1), but it's not recommended. 
