---
title: "Debugging methods in R6 objects"
output:
  html_document:
    theme: null
    css: mystyle.css
    toc: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Debugging methods in R6 objects}
  %\usepackage[utf8]{inputenc}
---

```{r echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Debugging methods in R6 classes is somewhat different from debugging normal R functions.

RStudio breakpoints don't work in R6 class methods. The simplest way to debug code is to insert a `browser()` line where you want to open a debugging console, reload the classes, and then step through your code. But this involves modifying your code, reloading it, and re-instantiating any objects you want to test.

## Enabling debugging for all future instances of a class

R6 generator objects have a method called `debug()` which will enable debugging for a method. This will affect all instances of the class that are created after the `debug()` is called.

```{r eval=FALSE}
# An example class
Simple <- R6Class("Simple",
  public = list(
    x = 10,
    getx = function() self$x
  )
)

# This will enable debugging the getx() method for objects of the 'Simple'
# class that are instantiated in the future.
Simple$debug("getx")

s <- Simple$new()
s$getx()
# [Debugging prompt]
```

To disable debugging for future instances, use the generator's `undebug()` method:

```{r eval=FALSE}
# Disable debugging for future instances:
Simple$undebug("getx")

s <- Simple$new()
s$getx()
#> [1] 10
```


## Debugging methods in individual objects

To enable debugging for a method in a single instance of an object, use the `debug()` function (not the `debug()` method in the generator object).

```{r eval=FALSE}
s <- Simple$new()
debug(s$getx)
s$getx()
# [Debugging prompt]
```

Use `undebug()` to disable debugging on an object's method.

```{r eval=FALSE}
undebug(s$getx)
s$getx()
#> [1] 10
```


You can also use the `trace()` function to specify where in a method you want to drop into the debugging console.
