---
title: "Speed of glue"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Speed of glue}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  % \VignetteDepends{R.utils
    R.utils,
    forcats,
    microbenchmark,
    rprintf,
    stringr,
    ggplot2}
---

Glue is advertised as

> Fast, dependency free string literals

So what do we mean when we say that glue is fast. This does not mean glue is
the fastest thing to use in all cases, however for the features it provides we
can confidently say it is fast.

A good way to determine this is to compare it's speed of execution to some alternatives.

- `base::paste0()`, `base::sprintf()` - Functions in base R implemented in C
  that provide variable insertion (but not interpolation).
- `R.utils::gstring()`, `stringr::str_interp()` - Provides a similar interface
  as glue, but using `${}` to delimit blocks to interpolate.
- `pystr::pystr_format()`[^1], `rprintf::rprintf()` - Provide a interfaces similar
  to python string formatters with variable replacement, but not arbitrary
  interpolation.

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = as.logical(Sys.getenv("EVAL_VIGNETTES", "FALSE")),
  cache = FALSE)
library(glue)
```

```{r setup2, include = FALSE}
plot_comparison <- function(x, ...) {
  library(ggplot2)
  library(microbenchmark)
  x$expr <- forcats::fct_reorder(x$expr, x$time)
  colors <- ifelse(levels(x$expr) == "glue", "orange", "grey")
  autoplot(x, ...) +
    theme(axis.text.y = element_text(color = colors)) +
      aes(fill = expr) + scale_fill_manual(values = colors, guide = FALSE)
}
```

## Simple concatenation

```{r, message = FALSE}
bar <- "baz"

simple <-
  microbenchmark::microbenchmark(
  glue = glue::glue("foo{bar}"),
  gstring = R.utils::gstring("foo${bar}"),
  paste0 = paste0("foo", bar),
  sprintf = sprintf("foo%s", bar),
  str_interp = stringr::str_interp("foo${bar}"),
  rprintf = rprintf::rprintf("foo$bar", bar = bar)
)

print(unit = "eps", order = "median", signif = 4, simple)

plot_comparison(simple)
```

While `glue()` is slower than `paste0`,`sprintf()` it is
twice as fast as `str_interp()` and `gstring()`, and on par with `rprintf()`.

`paste0()`, `sprintf()` don't do string interpolation and will likely always be
significantly faster than glue, glue was never meant to be a direct replacement
for them.

`rprintf()` does only variable interpolation, not arbitrary expressions, which
was one of the explicit goals of writing glue.

So glue is ~2x as fast as the two functions (`str_interp()`, `gstring()`) which do have
roughly equivalent functionality.

It also is still quite fast, with over 6000 evaluations per second on this machine.

## Vectorized performance

Taking advantage of glue's vectorization is the best way to avoid performance.
For instance the vectorized form of the previous benchmark is able to generate
100,000 strings in only 22ms with performance much closer to that of
`paste0()` and `sprintf()`. NB. `str_interp()` does not support
vectorization, so were removed.

```{r, message = FALSE}
bar <- rep("bar", 1e5)

vectorized <-
  microbenchmark::microbenchmark(
  glue = glue::glue("foo{bar}"),
  gstring = R.utils::gstring("foo${bar}"),
  paste0 = paste0("foo", bar),
  sprintf = sprintf("foo%s", bar),
  rprintf = rprintf::rprintf("foo$bar", bar = bar)
)

print(unit = "ms", order = "median", signif = 4, vectorized)

plot_comparison(vectorized, log = FALSE)
```

[^1]: pystr is no longer available from CRAN due to failure to correct
installation errors and was therefore removed from further testing.
